On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2019-05-08 manuscripts/IEEE/final_submission.tex
went over discussion

\begin{problem}\label{OptimizationProblem}
Find real $\mathbf{B}$ and \mbox{$\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$} that solve:
\begin{align}\label{minsum}
\min \sum_{\ell = 1}^N \|\mathbf{\overline x}_{\ell}\|_0 \ \
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta, \ \text{for all $i$}.
\end{align}
\end{problem}

Our development of Thm.~\ref{DeterministicUniquenessTheorem} is general enough to provide uniqueness stability even when generating $A$ do not fully satisfy (\ref{SparkCondition}) and recovery dictionaries $B$ have more columns than $A$.  Moreover, our approach incorporates a combinatorial theory for designing generating codes that should be of independent interest.
We also give brief arguments adapting our results to dictionaries and codes drawn from probability distributions (Cor.~\ref{ProbabilisticCor}). We defer to Sec.~\ref{DUT} the technical proofs of Thms.~\ref{DeterministicUniquenessTheorem} and ~\ref{SLCopt}, following some necessary definitions and a useful fact in combinatorial matrix analysis (Lem.~\ref{MainLemma}; proved in the Appendix). Finally, we discuss in Sec.~\ref{Discussion} theoretical and practical applications of our mathematical observations. 
%The Appendix contains a proof of Lem.~\ref{MainLemma}.

@@

Every $n \times \overline m$ matrix $\mathbf{B}$ for which there are $k$-sparse $\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$ satisfying \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \varepsilon$} for all $i \in [N]$ has $\overline m \geq m$, and provided $\overline m (r-1) < mr$, there is a permutation matrix $\mathbf{P}$ and an invertible diagonal matrix $\mathbf{D}$ such that:
\begin{align}\label{Cstable}
\|\mathbf{A}_j- \mathbf{BPD}_j\|_2 \leq C_1 \varepsilon, \ \ \text{for all } j \in J,
\end{align}
%
for some $J \subseteq [m]$ of size \mbox{$m - (r-1)(\overline m - m)$}. 
% $\overline m - r(\overline m - m)$

Moreover, if $\mathbf{A}$ satisfies \eqref{SparkCondition} and $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and for all $i \in [N]$:
\begin{align}\label{b-PDa}
%\|\mathbf{x}^J_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}^{\overline J}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}^{J}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon \ \  \text{for $i \in [N]$}.
\|(\mathbf{x}_i)_J - (\mathbf{D}^{-1}\mathbf{P}^{\top} \mathbf{\overline x}_i)_J\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,
\end{align}
%
where $(\mathbf{x}_i)_J$ represents the subvector formed from restricting to coefficients indexed by $J$. 
\end{theorem}

@@

A motivation for this work was the emergence of seemingly characteristic representations from sparse coding models fit to natural data, despite the varied assumptions underlying the many algorithms in current use. To this end, we have taken an important step toward unifying the many publications on the topic by demonstrating general sufficient, deterministic conditions under which identification of parameters in this model is not only possible but also robust to uncertainty in measurement and model choice.

We have shown that, given sufficient data, the problem of seeking a dictionary and sparse codes with minimal average support size (Prob.~\ref{OptimizationProblem}) reduces to an instance of Prob.~\ref{InverseProblem}, to which our main result (Thm.~\ref{DeterministicUniquenessTheorem}) applies: every dictionary and sequence of sparse codes consistent with the data are equivalent up to inherent relabeling/scaling ambiguities and a discrepancy (error) that scales linearly with the measurement noise or modeling inaccuracy. The constants we provide are explicit and computable; as such, there is an effective procedure that sufficiently affirms if a proposed solution to these problems is indeed unique up to noise and inherent ambiguities, although this computation is not efficient in general.

Our theoretical work mathematically justifies one of the few hypothesized theories of bottleneck communication in the brain \cite{Isely10}: that sparse neural population activity is recoverable from its noisy linear compression through a randomly constructed (but unknown) wiring bottleneck by any biologically plausible unsupervised sparse coding method that solves Prob.~\ref{DeterministicUniquenessTheorem} or \ref{SLCopt} (e.g., \cite{rehnsommer2007, rozell2007neurally, pehlevan2015normative}).\footnote{We refer the reader to \cite{ganguli2012compressed} for more on interactions between dictionary learning and neuroscience.}

@@

Beyond an original extension of existing noiseless guarantees \cite{Hillar15} to the noisy regime and their application to Prob.~\ref{OptimizationProblem}, a major innovation in our work is a theory of combinatorial designs for support sets key to the identification of the dictionary. We incorporate this idea into a fundamental lemma in matrix theory (Lem.~\ref{MainLemma}) that draws upon the definition of a new matrix lower bound induced by a hypergraph. Novel insights enabled by our combinatorial approach include: 1) a subset of dictionary elements is recoverable even if dictionary size is overestimated, 2) data require only a polynomial number of distinct sparse supports, and 3) recoverable dictionaries need not necessarily satisfy the spark condition.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and decode signals (e.g., in a device that learns structure from motion \cite{kong2016prior}), just as scientists have used it to make sense of their data \cite{jung2001imaging, Agarwal14, lee2016sparse, wu2016stability}. Assurances such as those offered by our theorems certify that different devices (with different initialization, etc.) will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these models of data; moreover, even small noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, it seems a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis is the assumed well-posedness of parameter identification in the sparse coding model, confirmation of which forms the core of our theoretical findings.

>>>>

\begin{problem}\label{OptimizationProblem}
Find real $\mathbf{B}$ and \mbox{$\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$} that solve:
\begin{align}\label{minsum}
\min \sum_{i = 1}^N \|\mathbf{\overline x}_{i}\|_0 \ \
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta, \ \text{for all $i$}.
\end{align}
\end{problem}

Our development of Thm.~\ref{DeterministicUniquenessTheorem} is general enough to provide some uniqueness stability even when generating $A$ do not fully satisfy (\ref{SparkCondition}) and recovery dictionaries $B$ have more columns than $A$.  Moreover, the approach incorporates a combinatorial theory for designing generating codes that should be of independent interest. We also give brief arguments adapting our results to dictionaries and codes drawn from probability distributions (Cor.~\ref{ProbabilisticCor}). We defer to Sec.~\ref{DUT} the technical proofs of Thms.~\ref{DeterministicUniquenessTheorem} and ~\ref{SLCopt}, following some necessary definitions and a useful fact in combinatorial matrix analysis (Lem.~\ref{MainLemma}; proved in the Appendix). Finally, we discuss in Sec.~\ref{Discussion} theoretical and practical applications of our mathematical observations. 
%The Appendix contains a proof of Lem.~\ref{MainLemma}.

@@

Every $n \times \overline m$ matrix $\mathbf{B}$ for which there are $k$-sparse $\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$ satisfying \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \varepsilon$} for all $i \in [N]$ has $\overline m \geq m$, and provided $\overline m (r-1) < mr$, there is a permutation matrix $\mathbf{P}$ and an invertible diagonal matrix $\mathbf{D}$ such that:
\begin{align}\label{Cstable}
\|\mathbf{A}_j- (\mathbf{BPD})_j\|_2 \leq C_1 \varepsilon, \ \ \text{for all } j \in J,
\end{align}
%
for some $J \subseteq [m]$ of size \mbox{$m - (r-1)(\overline m - m)$}. 
% $\overline m - r(\overline m - m)$

Moreover, if $\mathbf{A}$ satisfies \eqref{SparkCondition} and $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and for all $i \in [N]$:
\begin{align}\label{b-PDa}
%\|\mathbf{x}^J_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}^{\overline J}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}^{J}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon \ \  \text{for $i \in [N]$}.
\|(\mathbf{x}_i)_J - (\mathbf{D}^{-1}\mathbf{P}^{\top} \mathbf{\overline x}_i)_J\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,
\end{align}
%
where $(\mathbf{x}_i)_J$ represents the subvector formed from restricting to coordinates indexed by $J$. 
\end{theorem}

@@

A main motivation for this work is the emergence of seemingly unique representations from sparse coding models trained on natural data, despite the varied assumptions underlying the many algorithms in current use. Our results constitute an important step toward explaining these phenomena as well as unifying many publications on the topic by demonstrating general deterministic conditions under which identification of parameters in this model is not only possible but also robust to uncertainty in measurement and model choice.

We have shown that, given sufficient data, the problem of determining a dictionary and sparse codes with minimal support size (Prob.~\ref{OptimizationProblem}) reduces to an instance of Prob.~\ref{InverseProblem}, to which our main result (Thm.~\ref{DeterministicUniquenessTheorem}) applies: every dictionary and sequence of sparse codes consistent with the data are equivalent up to inherent relabeling/scaling ambiguities and a discrepancy (error) that scales linearly with the measurement noise or modeling inaccuracy. The constants we provide are explicit and computable; as such, there is an effective procedure that sufficiently affirms if a proposed solution to these problems is indeed unique up to noise and inherent ambiguities, although it is not efficient in general.

One direct application of our work is to neural theory, wherein our theorems justify the mathematical soundness of one of the few hypothesized theories of bottleneck communication in the brain \cite{Isely10}: that sparse neural population activity is recoverable from its noisy linear compression through a randomly constructed (but unknown) wiring bottleneck by any biologically plausible unsupervised sparse coding method that solves Prob.~\ref{DeterministicUniquenessTheorem} or \ref{SLCopt} (e.g., \cite{rehnsommer2007, rozell2007neurally, pehlevan2015normative}).\footnote{We refer the reader to \cite{ganguli2012compressed} for more on interactions between dictionary learning and neuroscience.}

@@

Beyond an extension of existing noiseless guarantees \cite{Hillar15} to the noisy regime and their novel application to Prob.~\ref{OptimizationProblem}, our work contains a theory of combinatorial designs for support sets key to the identification of the dictionary. We incorporate this idea into a fundamental lemma in matrix theory (Lem.~\ref{MainLemma}) that draws upon the definition of a new matrix lower bound induced by a hypergraph. New insights enabled by this combinatorial approach include: 1) a subset of dictionary elements is recoverable even if dictionary size is overestimated, 2) data require only a polynomial number of distinct sparse supports, and 3) recoverable dictionaries need not necessarily satisfy the spark condition.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks; e.g., for MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and even the design of an ultrafast camera \cite{Gao14}. It is only a matter of time before these systems incorporate dictionary learning to encode and decode signals (e.g., in a device that learns structure from motion \cite{kong2016prior}), just as scientists have used it to make sense of their data \cite{jung2001imaging, Agarwal14, lee2016sparse, wu2016stability}.

Assurances such as those offered by our theorems certify that different devices and algorithms (with different initialization, etc.) will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these models of data; moreover, even small noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis seems to be the assumed well-posedness of parameter identification in the sparse coding model, confirmation of which forms the core of our theoretical findings.
